### 2022-06-27 main/chapter_2ljets.tex
smash out some matrix method theory and opinions

@@

>>>>

% TODO jetM
\TODO{JET MASS}

@@

\TODO{composition (INT)}
% components
% lots of non-prompt, a little fake hadrons
%

Historical precedent leads us to use the `Matrix Method' to estimate
backgrounds including fake and non-prompt leptons, so named because it works by
estimating a linear relationship between dilepton event rates.
\TODO{previous 2ljets MM}

In data, one measures tightly ($t$) and loosely ($l$) identified leptons, leading
to four categories of events with two ordered leptons as labelled $tt$, $tl$,
$lt$, and $ll$ for tight-tight, tight-loose, and so on.
The Matrix Method assumes a linear relationship between these selections
and those of real ($r$) and fake/non-prompt ($f$) leptons:
\begin{equation}
\begin{pmatrix}
\mu^\textrm{data}_{tt} \\
\mu^\textrm{data}_{tl} \\
\mu^\textrm{data}_{tl} \\
\mu^\textrm{data}_{ll} \\
\end{pmatrix}
=
M
\begin{pmatrix}
\mu^\textrm{truth}_{rr} \\
\mu^\textrm{truth}_{fr} \\
\mu^\textrm{truth}_{rf} \\
\mu^\textrm{truth}_{ff} \\
\end{pmatrix}
,
\end{equation}
where $\mu^{\ldots}_{**}$ is a Poisson event rate and $M$ is the matrix of
transition probabilities after which this method is
named~\cite{ATLAS-CONF-2014-058}.

Each element of $M$ is a probability for a truth-level (real/fake) event to be
also in a corresponding data-level (tight/loose) category.
For example, its top right element $M_{{tt}, {rr}}$ is the
probability that an event with two real leptons has them both be identified
tightly.
The leptons are further assumed to be reconstructed independently, such that
$M_{{tt}, {rr}} =
p(t\!\mid\!r,1)
\times p(t\!\mid\!r,2)$
in which the numerical index points to the first and second leptons
respectively.

Transition probabilities $p(*\!\mid\!*,i)$ are estimated by studies in
separate selections and assigned as functions of the lepton
flavour, $\pt$ and $|\eta|$.
Given those estimated transition probabilities and some observed data
in a selection of two loose leptons,
$n^\textrm{data}_{tt}$,
$n^\textrm{data}_{tl}$,
$n^\textrm{data}_{tl}$,
and $n^\textrm{data}_{ll}$,
the Matrix Method inverts $M$ to estimate truth level rates
$\bar \mu^\textrm{truth}_{**}$.
(Recall that tight implies loose, so this selection of loose leptons includes
some tight leptons.)
Finally, these truth-level estimates are converted back to data-level
estimates in the tight-tight selection of our search regions back through the
independent transition probabilities as
\begin{align}
% rr
\bar \mu^\textrm{data}_{tt\mid rr} &=
p(t\!\mid\!r,1)
\times p(t\!\mid\!r,2)
\times \bar \mu^\textrm{truth}_{rr}~~, \\
% fr
\bar \mu^\textrm{data}_{tt\mid fr} &=
p(t\!\mid\!f,1)
\times p(t\!\mid\!r,2)
\times \bar \mu^\textrm{truth}_{fr}~~, \\
% rf
\bar \mu^\textrm{data}_{tt\mid rf} &=
p(t\!\mid\!r,1)
\times p(t\!\mid\!f,2)
\times \bar \mu^\textrm{truth}_{rf}~~,~~\textrm{and} \\
% ff
\bar \mu^\textrm{data}_{tt\mid ff} &=
p(t\!\mid\!f,1)
\times p(t\!\mid\!f,2)
\times \bar \mu^\textrm{truth}_{ff}~~.
\end{align}

To assign fake/non-prompt estimates in our regions, this method is
implemented by assigning weights to observed data events, which are
histogrammed in the tighter selections of our search
regions~\cite{twoljets2018int}.
Systematic variations are assigned by comparing histograms when those weights
are varied according to various uncertainties in the Matrix Method procedure.

Yes, that is a histogramming of real data with weights, and those data include
tight-tight events which comprise the signal data of our search regions,
in which we are searching for anomalies.
Since it includes real data, this method looks at all unblinded data, including
signal regions.
And because it knows the data we are supposed to be predicting, it is no
prediction at all!

That last statement was an exaggeration.
Although it does cheat by using the data we should be blind to, it uses those
data in a diluted way, such that they which might end up not making much
difference.

In fact, real data tend to be assigned negative Matrix Method weights.
For example, CR-VZ receives $157$ weighted data in its Matrix Method sample,
of which $135$ pass the signal lepton requirements (tight-tight)%
\footnote{%
I am not sure why this differs from the $194$ actual data observed in CR-VZ.
The two do overlap in event identity numbers, so there is presumably some
tighter selection in the Matrix Method which should be accounted for in its
weights.
},
and the remainder have one loose lepton (zero loose-loose lepton events
are observed here).
The signal events have an average weight of $-0.0227$, and the others
have an average weight of $+0.238$ (one can check these recover the central
fake/non-prompt background of $2.2$).
Such negative signal weights suggest that a local \emph{excess} of data
would act to \emph{decrease} the Matrix Method estimate in the region.
Yes this backwards dependency is backwards, but this gives some relief that it
should not hide real local excesses.

Although the $\twoljets$-electroweak regions were designed without using the
Matrix Method fake estimate, and before it was finalized, we are fortunate that
most search regions do not have substantial Matrix Method yields.
I consider those that do to be dubious.

% TODO cite past searches
% TODO thank Eirik
